Automated assessment of residual plots with computer vision models

### 2024-04-15 paper/paper.tex
update paper

% Pandoc syntax highlighting
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

\hypertarget{methodology}{%
\section{Methodology}\label{methodology}}

\hypertarget{different-possible-configurations-of-the-model-formula}{%
\subsection{Different possible configurations of the model
formula}\label{different-possible-configurations-of-the-model-formula}}

@@

\hypertarget{input-formats}{%
\subsubsection{Input formats}\label{input-formats}}

@@

\hypertarget{output-formats}{%
\subsubsection{Output formats}\label{output-formats}}

@@

\hypertarget{distance-from-the-good-residual-plots}{%
\subsection{Distance from the good residual
plots}\label{distance-from-the-good-residual-plots}}

@@

\hypertarget{residual-distribution}{%
\subsubsection{Residual distribution}\label{residual-distribution}}

@@

\hypertarget{kullback-leibler-divergence-of-p-from-q}{%
\subsubsection{\texorpdfstring{Kullback-Leibler divergence of \(P\) from
\(Q\)}{Kullback-Leibler divergence of P from Q}}\label{kullback-leibler-divergence-of-p-from-q}}

@@

\hypertarget{evaluation-of-kullback-leibler-divergence-for-non-normal-p}{%
\subsubsection{\texorpdfstring{Evaluation of Kullback-Leibler divergence
for non-normal
\(P\)}{Evaluation of Kullback-Leibler divergence for non-normal P}}\label{evaluation-of-kullback-leibler-divergence-for-non-normal-p}}

@@

\hypertarget{approximation-of-the-distance-measure}{%
\subsubsection{Approximation of the distance
measure}\label{approximation-of-the-distance-measure}}

@@

\hypertarget{approximated-distance-as-an-model-violations-index-mvi}{%
\subsubsection{Approximated distance as an Model Violations Index
(MVI)}\label{approximated-distance-as-an-model-violations-index-mvi}}

@@

Note that since the Kullback-Leibler divergence could be affected by the
number of random variables involved in the evaluation. In general, more
observations will result in greater distance \(D\). This does not mean
\(\hat{D}\) can not correctly represent the extent of the model
violations since more observations in a residual plot usually lead to
stronger visual signal strength as the underlying visual patterns will
have higher chance to be revealed except for cases like serious
overlapping. For the same data generating process, \(D\) usually grows
logarithmically as the number of observations increases. This is because
\(D = \text{log}(1 + D_{KL})\), and
\(D_{KL} = \sum_{i=1}^{n}D_{KL}^{(i)}\) under the independence
assumption.

@@

\hypertarget{statistical-testing-based-on-the-approximated-distance}{%
\subsection{Statistical testing based on the approximated
distance}\label{statistical-testing-based-on-the-approximated-distance}}

\hypertarget{null-distribution-of-the-approximated-distance}{%
\subsubsection{Null distribution of the approximated
distance}\label{null-distribution-of-the-approximated-distance}}

@@

To solve this issue while aligning with the principle of visual
inference, \(\hat{D}\) can be viewed as a test statistic. And the null
distribution of this statistic can be approximated by the empirical
distribution of \(\hat{D}_{null}\), where \(\hat{D}_{null}\) is the
approximated distance for a null plot simulated from the fitted
regression model. The approximation of the distribution involves
applying the residual rotation technique \citep{buja2009statistical} on
the fitted regression model to obtain null residuals. The null residuals
are then used to make null plots and fed into the computer vision model
to get predictions. The empirical distribution is constructed with the
approximated distance.

There are two types of error associated with the approximation of the
null distribution. These include the sampling error and the estimation
error for model parameters. Increasing the number of null plots within
the distribution approximation can effectively reduce sampling error.
However, the estimation error for model parameters remains irreducible.
The residual rotation technique operates under the assumption of a
correct fitted model, using the estimated variance instead of the
unknown true variance to rescale rotated residuals. Consequently, under
\(H_0\), the true distribution of \(\hat{D}_{null}\) will slightly
deviate from the approximated null distribution.

\hypertarget{estimation-of-quantiles-of-the-null-distribution}{%
\subsubsection{Estimation of quantiles of the null
distribution}\label{estimation-of-quantiles-of-the-null-distribution}}

Let \(\hat{D}_{null}^{(i)}\) be the approximated distance of the
\(i\)-th null plots, where \(i = 1,...,n_{null}\) and
\(n_{null} \in \mathbb{N}^+\) is a sufficiently large number. Quantiles
of the null distribution can then be estimated using the sample
quantiles available in statistical software such as R. The details of
the sample quantile computation can be found in
\citet{hyndman1996sample}. In statistical testing, analysts often care
about certain quantiles of the null distribution, such as the 90\%
quantile, the 95\% quantile and the 99\% quantile. These quantiles are
used as critical values to decide if \(H_0\) needs to be rejected. For
example, if \(\hat{D}\) is greater than and equal to the 95\% sample
quantile \(Q_{null}(0.95)\), we could say the approximated distance for
the actual residual plot is significantly different from the
approximated distance for null plots with 95\% significance level. Based
on our experience, in order to get a stable estimate of the 95\%
quantile, \(n_{null}\) usually needs to be at least \(100\). And if the
null distribution has a long tail, more null plots will be needed.
Alternatively, a p-value can be used to represents the probability of
observing an event equally or more extreme than the given event under
\(H_0\), and it can be estimated by
\(1/m\sum_{i=1}^{n_{null}}I\left(\hat{D}_{null}^{(i)} \geq \hat{D}\right)\).

@@

\hypertarget{bootstrapping-the-approximated-distance}{%
\subsubsection{Bootstrapping the approximated
distance}\label{bootstrapping-the-approximated-distance}}

@@

\hypertarget{generation-of-training-data}{%
\subsection{Generation of training
data}\label{generation-of-training-data}}

\hypertarget{data-generating-process}{%
\subsubsection{Data generating process}\label{data-generating-process}}

@@

\hypertarget{computation-of-scagnostics}{%
\subsubsection{Computation of
scagnostics}\label{computation-of-scagnostics}}

@@

\hypertarget{crafting-a-balanced-training-set}{%
\subsubsection{Crafting a balanced training
set}\label{crafting-a-balanced-training-set}}

@@

\hypertarget{architecture-of-the-computer-vision-model}{%
\subsection{Architecture of the computer vision
model}\label{architecture-of-the-computer-vision-model}}

@@

\hypertarget{training-and-hyperparameter-tuning}{%
\subsection{Training and hyperparameter
tuning}\label{training-and-hyperparameter-tuning}}

@@

\hypertarget{model-evaluation-methods}{%
\subsection{Model evaluation methods}\label{model-evaluation-methods}}

@@

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{optimized-hyperparameters}{%
\subsection{Optimized hyperparameters}\label{optimized-hyperparameters}}

@@

All optimized models chose to retain the additional inputs, contributing
to the reduction of validation error. The number of units required for
the fully-connected layer was \(256\), a relatively modest number
compared to the VGG16 classifier, suggesting that the problem at hand
was relatively straightforward. The optimized learning rates were
typically smaller than the default value recommended by Keras, which is
0.001.

@@

\hypertarget{model-performance}{%
\subsection{Model performance}\label{model-performance}}

@@

The regression tree revealed that most issues arose from non-linearity
problems and the presence of a second predictor in the regression model.
When the error distribution had a very small variance, all computer
vision models tended to under-predict the distance. Conversely, when the
error distribution had a large variance, all models tended to
over-predict the distance. Therefore, Additionally, for input images
representing null plots, it was expected that the models will
over-predict the distance, as explained in Section
\ref{null-distribution-of-the-approximated-distance}.

Since most of the deviation stemmed from the presence of non-linearity
violations, to further investigate this, we split the test set and
re-evaluated the performance, as detailed in Table
\ref{tab:performance-sub}. It was evident that metrics for null plots
were notably worse compared to other categories. Furthermore, residual
plots solely exhibiting non-normality issues were the easiest to
predict, with very low test root mean square error (RMSE) around
\(0.3\), very low mean absolute error (MAE) around \(0.2\), and very
high \(R^2\) around \(0.94\). Residual plots with non-linearity issues
were more challenging to assess than those with heteroskedasticity or
non-normality issues. Assessing residual plots with heteroskedasticity
was not as difficult as assessing those with non-linearity issues. When
multiple violations were introduced to a residual plot, the performance
metrics typically lay between the metrics for each individual violation.

Based on the model performance metrics, we will only use the
best-performing model evaluated on the test set, namely the
\(32 \times 32\) model, for the subsequent analysis.

@@

\caption{\label{tab:performance}The training and test performance of three optimized models with different input sizes. The best metrics are colored in red.}
\centering
\begin{tabular}[t]{l>{}l>{}l>{}l>{}l}
\toprule
 & RMSE & $R^2$ & MAE & Huber loss\\
\midrule
\addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Training set}}\\
\hspace{1em}$32 \times 32$ & \textcolor{black}{0.531} & \textcolor{black}{0.937} & \textcolor{black}{0.364} & \textcolor{black}{0.126}\\
\hspace{1em}$64 \times 64$ & \textcolor{red}{0.405} & \textcolor{red}{0.963} & \textcolor{red}{0.260} & \textcolor{red}{0.072}\\
\hspace{1em}$128 \times 128$ & \textcolor{black}{0.432} & \textcolor{black}{0.959} & \textcolor{black}{0.290} & \textcolor{black}{0.084}\\
\addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Test set}}\\
\hspace{1em}$32 \times 32$ & \textcolor{red}{0.660} & \textcolor{red}{0.901} & \textcolor{red}{0.434} & \textcolor{red}{0.181}\\
\hspace{1em}$64 \times 64$ & \textcolor{black}{0.674} & \textcolor{black}{0.897} & \textcolor{black}{0.438} & \textcolor{black}{0.186}\\
\hspace{1em}$128 \times 128$ & \textcolor{black}{0.692} & \textcolor{black}{0.892} & \textcolor{black}{0.460} & \textcolor{black}{0.199}\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:performance-sub}The training and test performance of the $32 \times 32$ model presented with different model violations.  The best metrics are colored in red.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{ll>{}l>{}l>{}l>{}l}
\toprule
Violations & \#samples & RMSE & $R^2$ & MAE & Huber loss\\
\midrule
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Training set}}\\
\hspace{1em}no violations & 1546 & \textcolor{black}{1.149} & \textcolor{black}{} & \textcolor{black}{1.080} & \textcolor{black}{0.592}\\
\hspace{1em}non-linearity & 22184 & \textcolor{black}{0.625} & \textcolor{black}{0.851} & \textcolor{black}{0.474} & \textcolor{black}{0.177}\\
\hspace{1em}heteroskedasticity & 10826 & \textcolor{black}{0.528} & \textcolor{black}{0.942} & \textcolor{black}{0.407} & \textcolor{black}{0.133}\\
\hspace{1em}non-linearity + heteroskedasticity & 10221 & \textcolor{black}{0.593} & \textcolor{black}{0.920} & \textcolor{black}{0.439} & \textcolor{black}{0.158}\\
\hspace{1em}non-normality & 10797 & \textcolor{red}{0.279} & \textcolor{red}{0.955} & \textcolor{red}{0.177} & \textcolor{red}{0.038}\\
\hspace{1em}non-linearity + non-normality & 8835 & \textcolor{black}{0.472} & \textcolor{black}{0.923} & \textcolor{black}{0.274} & \textcolor{black}{0.090}\\
\hspace{1em}heteroskedasticity + non-normality & 7870 & \textcolor{black}{0.354} & \textcolor{black}{0.905} & \textcolor{black}{0.211} & \textcolor{black}{0.052}\\
\hspace{1em}non-linearity + heteroskedasticity + non-normality & 7721 & \textcolor{black}{0.432} & \textcolor{black}{0.874} & \textcolor{black}{0.261} & \textcolor{black}{0.078}\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Test set}}\\
\hspace{1em}no violations & 155 & \textcolor{black}{1.267} & \textcolor{black}{} & \textcolor{black}{1.174} & \textcolor{black}{0.685}\\
\hspace{1em}non-linearity & 2218 & \textcolor{black}{0.787} & \textcolor{black}{0.771} & \textcolor{black}{0.578} & \textcolor{black}{0.257}\\
\hspace{1em}heteroskedasticity & 1067 & \textcolor{black}{0.602} & \textcolor{black}{0.922} & \textcolor{black}{0.461} & \textcolor{black}{0.170}\\
\hspace{1em}non-linearity + heteroskedasticity & 985 & \textcolor{black}{0.751} & \textcolor{black}{0.868} & \textcolor{black}{0.541} & \textcolor{black}{0.236}\\
\hspace{1em}non-normality & 1111 & \textcolor{red}{0.320} & \textcolor{red}{0.942} & \textcolor{red}{0.194} & \textcolor{red}{0.049}\\
\hspace{1em}non-linearity + non-normality & 928 & \textcolor{black}{0.600} & \textcolor{black}{0.879} & \textcolor{black}{0.335} & \textcolor{black}{0.133}\\
\hspace{1em}heteroskedasticity + non-normality & 819 & \textcolor{black}{0.489} & \textcolor{black}{0.832} & \textcolor{black}{0.267} & \textcolor{black}{0.093}\\
\hspace{1em}non-linearity + heteroskedasticity + non-normality & 717 & \textcolor{black}{0.620} & \textcolor{black}{0.762} & \textcolor{black}{0.339} & \textcolor{black}{0.140}\\
\bottomrule
\end{tabular}}
\end{table}

\hypertarget{model-violations-index-mvi}{%
\subsection{Model Violations Index
(MVI)}\label{model-violations-index-mvi}}

Figures \ref{fig:poly-index} and \ref{fig:heter-index} display the
residual plots for fitted models exhibiting varying degrees of
non-linearity and heteroskedasticity. Each residual plot's Model
Violation Index (MVI) is computed using Equation \ref{eq:mvi} with
\(C = 10\). When \(\text{MVI} > 8\), the visual patterns are notably
strong and easily discernible by humans. In the range
\(6 < \text{MVI} < 8\), the visibility of the visual pattern diminishes
as MVI decreases. Conversely, when \(\text{MVI} < 6\), the visual
pattern tends to become relatively faint and challenging to observe.
Table \ref{tab:mvi} provides a summary of the MVI usage and it is
applicable to other linear regression models.

@@

\hypertarget{comparison-with-human-visual-inference-and-conventional-tests}{%
\subsection{Comparison with human visual inference and conventional
tests}\label{comparison-with-human-visual-inference-and-conventional-tests}}

\hypertarget{overview-of-the-human-subject-experiment}{%
\subsubsection{Overview of the human subject
experiment}\label{overview-of-the-human-subject-experiment}}

@@

\hypertarget{model-performance-on-the-human-data}{%
\subsubsection{Model performance on the human
data}\label{model-performance-on-the-human-data}}

@@

For each lineup used in \citet{li2023plot}, there is one actual residual
plot and 19 null plots. While the distance \(D\) for the actual residual
plot depends on the underlying data generating process, the distance
\(D\) for the null plots is zero. We have used our optimized computer
vision model to predict \(\hat{D}\) for both the actual residual plots
and the null plots. Additionally, the appropriate conventional tests
including the Ramsey Regression Equation Specification Error Test
(RESET) \citep{ramsey1969tests} for non-linearity and the Breusch-Pagan
test \citep{breusch1979simple} for heteroskedasticity were applied on
the same data for comparison.

@@

For null plots, a histogram of the predictions is provided in Figure
\ref{fig:hist-null-human}. In most cases, \(\hat{D}\) is predicted to be
around \(1\). As \(\hat{D}\) increases, the number of predictions
decreases. The 90\% and 95\% sample quantiles are approximately at
\(2\), and the 99\% sample quantile is around \(2.5\). These sample
quantiles can serve as global critical values for determining whether a
residual plot needs to be rejected. We have also computed the 95\%
sample quantiles by assessing 200 null plots for each fitted model used
in the experiment. The difference between decisions based on these two
types of critical values is minor, with decisions aligning 93.8\% of the
time. Therefore, we will be using the 95\% sample quantiles obtained
from each fitted model in the following analysis.

>>>>

% Pandoc syntax highlighting
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

@@

\section{Introduction}\label{introduction}

@@

\section{Methodology}\label{methodology}

\subsection{Different possible configurations of the model
formula}\label{different-possible-configurations-of-the-model-formula}

@@

\subsubsection{Input formats}\label{input-formats}

@@

\subsubsection{Output formats}\label{output-formats}

@@

\subsection{Distance from the good residual
plots}\label{distance-from-the-good-residual-plots}

@@

\subsubsection{Residual distribution}\label{residual-distribution}

@@

\subsubsection{\texorpdfstring{Kullback-Leibler divergence of \(P\) from
\(Q\)}{Kullback-Leibler divergence of P from Q}}\label{kullback-leibler-divergence-of-p-from-q}

@@

\subsubsection{\texorpdfstring{Evaluation of Kullback-Leibler divergence
for non-normal
\(P\)}{Evaluation of Kullback-Leibler divergence for non-normal P}}\label{evaluation-of-kullback-leibler-divergence-for-non-normal-p}

@@

\subsubsection{Approximation of the distance
measure}\label{approximation-of-the-distance-measure}

@@

\subsubsection{Model Violations Index
(MVI)}\label{sec:model-violations-index}

@@

@@

\subsection{Statistical testing based on the approximated
distance}\label{statistical-testing-based-on-the-approximated-distance}

\subsubsection{Lineup evaluation}\label{lineup-evaluation}

@@

To address this issue while adhering to the principle of visual
inference, we can compare the estimated distance \(\hat{D}\) to the
approximated distances for the null plots in a lineup. Specifically, if
a lineup comprises 20 plots, the null hypothesis \(H_0\) will be
rejected if \(\hat{D}\) exceeds the maximum approximated distance among
the \(m - 1\) null plots, denoted as
\(\max\limits_{1 \leq i \leq m-1} {\hat{D}_{null}^{(i)}}\), where
\(\hat{D}_{null}^{(i)}\) represents the approximated distance for the
\(i\)-th null plot. This approach is equivalent to the typical lineup
protocol requiring a 95\% significance level, where \(H_0\) is rejected
if the data plot is identified as the most distinct plot by the sole
participant. The approximated distance serves as a metric to quantify
the difference between the data plot and the null plots, as intended.

For lineups consisting of more than 20 plots, the 95\% significance
level can be maintained if the number of plots is a multiple of 20.
Specifically, for lineups comprising \(20p\) plots, where \(p\) is a
positive integer, we reject \(H_0\) if \(\hat{D}\) exceeds 95\%
\({\hat{D}_{null}^{(i)}}\) for \(i = 1, \ldots, 20p-1\). The p-value in
this case is given by
\(\frac{1}{m} + \frac{1}{m}\sum{i=1}^{m-1} I\left(\hat{D}_{null}^{(i)} > \hat{D}\right)\),
where \(I(\cdot)\) is the indicator function.

Moreover, if the number of plots in a lineup, denoted by \(m\), is
sufficiently large, the empirical distribution of
\({\hat{D}_{null}^{(i)}}\) can be viewed as an approximation of the null
distribution of the approximated distance. Consequently, quantiles of
the null distribution can be estimated using the sample quantiles
available in statistical software such as R, and these quantiles can be
utilized for decision-making purposes. The details of the sample
quantile computation can be found in \citet{hyndman1996sample}. For
instance, if \(\hat{D}\) is greater than or equal to the 95\% sample
quantile, denoted as \(Q_{null}(0.95)\), we can conclude that the
approximated distance for the actual residual plot is significantly
different from the approximated distance for null plots with a 95\%
significance level. Based on our experience, to obtain a stable estimate
of the 95\% quantile, the number of null plots, \(n_{null}\), typically
needs to be at least 100. However, if the null distribution exhibits a
long tail, a larger number of null plots may be required. Alternatively,
a p-value can be used to represent the probability of observing an event
equally or more extreme than the given event under the null hypothesis
\(H_0\), and it can be estimated by
\(\frac{1}{m}\sum_{i=1}^{m-1}I\left(\hat{D}_{null}^{(i)} \geq \hat{D}\right)\).

@@

\subsubsection{Bootstrapping the approximated
distance}\label{bootstrapping-the-approximated-distance}

@@

\subsection{Generation of training
data}\label{generation-of-training-data}

\subsubsection{Data generating process}\label{data-generating-process}

@@

\subsubsection{Computation of
scagnostics}\label{computation-of-scagnostics}

@@

\subsubsection{Crafting a balanced training
set}\label{crafting-a-balanced-training-set}

@@

\subsection{Architecture of the computer vision
model}\label{architecture-of-the-computer-vision-model}

@@

\subsection{Training and hyperparameter
tuning}\label{training-and-hyperparameter-tuning}

@@

\subsection{Model evaluation methods}\label{model-evaluation-methods}

@@

\section{Results}\label{results}

\subsection{Optimized hyperparameters}\label{optimized-hyperparameters}

@@

All optimized models chose to retain the additional inputs, contributing
to the reduction of validation error. The number of units required for
the fully-connected layer was \(256\), a relatively modest number
compared to the VGG16 classifier, suggesting that the problem at hand
was less complex. The optimized learning rates were smaller than the
default value recommended by Keras, which is 0.001.

@@

\subsection{Summary of model
performance}\label{summary-of-model-performance}

@@

The regression tree revealed that most issues arose from non-linearity
problems and the presence of a second predictor in the regression model.
When the error distribution had a very small variance, all computer
vision models tended to under-predict the distance. Conversely, when the
error distribution had a large variance, all models tended to
over-predict the distance. Furthermore, for input images representing
null plots, it was expected that the models will over-predict the
distance, as explained in Section \ref{lineup-evaluation}.

Since most of the deviation stemmed from the presence of non-linearity
violations, to further investigate this, we split the test set and
re-evaluated the performance, as detailed in Table
\ref{tab:performance-sub}. It was evident that metrics for null plots
were notably worse compared to other categories. Furthermore, residual
plots solely exhibiting non-normality issues were the easiest to
predict, with very low test root mean square error (RMSE) at around
\(0.3\). Residual plots with non-linearity issues were more challenging
to assess than those with heteroskedasticity or non-normality issues.
Assessing residual plots with heteroskedasticity was not as difficult as
assessing those with non-linearity issues. When multiple violations were
introduced to a residual plot, the performance metrics typically lay
between the metrics for each individual violation.

Based on the model performance metrics, we chose to use the
best-performing model evaluated on the test set, namely the
\(32 \times 32\) model, for the subsequent analysis.

@@

\caption{\label{tab:performance}The training and test performance of three optimized models with different input sizes.}
\centering
\begin{tabular}[t]{lllll}
\toprule
 & RMSE & $R^2$ & MAE & Huber loss\\
\midrule
\addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Training set}}\\
\hspace{1em}$32 \times 32$ & 0.531 & 0.937 & 0.364 & 0.126\\
\hspace{1em}$64 \times 64$ & 0.405 & 0.963 & 0.260 & 0.072\\
\hspace{1em}$128 \times 128$ & 0.432 & 0.959 & 0.290 & 0.084\\
\addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Test set}}\\
\hspace{1em}$32 \times 32$ & 0.660 & 0.901 & 0.434 & 0.181\\
\hspace{1em}$64 \times 64$ & 0.674 & 0.897 & 0.438 & 0.186\\
\hspace{1em}$128 \times 128$ & 0.692 & 0.892 & 0.460 & 0.199\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:performance-sub}The training and test performance of the $32 \times 32$ model presented with different model violations.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lllll}
\toprule
Violations & \#training samples & Training RMSE & \#test samples & Test RMSE\\
\midrule
no violations & 1546 & 1.149 & 155 & 1.267\\
non-linearity & 22184 & 0.625 & 2218 & 0.787\\
heteroskedasticity & 10826 & 0.528 & 1067 & 0.602\\
non-linearity + heteroskedasticity & 10221 & 0.593 & 985 & 0.751\\
non-normality & 10797 & 0.279 & 1111 & 0.320\\
\addlinespace
non-linearity + non-normality & 8835 & 0.472 & 928 & 0.600\\
heteroskedasticity + non-normality & 7870 & 0.354 & 819 & 0.489\\
non-linearity + heteroskedasticity + non-normality & 7721 & 0.432 & 717 & 0.620\\
\bottomrule
\end{tabular}}
\end{table}

\subsection{Demostration of MVI}\label{demostration-of-mvi}

As described in Section \ref{sec:model-violations-index} a Model
Violations Index (MVI) can be constructed using \(\hat{D}\). This
section provides guidelines and examples of MVI.

Figures \ref{fig:poly-index} and \ref{fig:heter-index} display the
residual plots for fitted models exhibiting varying degrees of
non-linearity and heteroskedasticity. Each residual plot's MVI is
computed using Equation \ref{eq:mvi} with \(C = 10\). When
\(\text{MVI} > 8\), the visual patterns are notably strong and easily
discernible by humans. In the range \(6 < \text{MVI} < 8\), the
visibility of the visual pattern diminishes as MVI decreases.
Conversely, when \(\text{MVI} < 6\), the visual pattern tends to become
relatively faint and challenging to observe. Table \ref{tab:mvi}
provides a summary of the MVI usage and it is applicable to other linear
regression models.

@@

\subsection{Comparison with human visual inference and conventional
tests}\label{comparison-with-human-visual-inference-and-conventional-tests}

\subsubsection{Overview of the human subject
experiment}\label{overview-of-the-human-subject-experiment}

@@

\subsubsection{Model performance on the human
data}\label{model-performance-on-the-human-data}

@@

For each lineup used in \citet{li2023plot}, there is one actual residual
plot and 19 null plots. While the distance \(D\) for the actual residual
plot depends on the underlying data generating process, the distance
\(D\) for the null plots is zero. We have used our optimized computer
vision model to predict the approximated distance for both the actual
residual plots and the null plots. To have a fair comparison, \(H_0\)
will be rejected if the actual residual plot has the greatest
approximated distance among all plots in a lineup. Additionally, the
appropriate conventional tests including the Ramsey Regression Equation
Specification Error Test (RESET) \citep{ramsey1969tests} for
non-linearity and the Breusch-Pagan test \citep{breusch1979simple} for
heteroskedasticity were applied on the same data for comparison.

@@

Table \ref{tab:human-conv-table} provides a summary of the agreement
between decisions made by the computer vision model and conventional
tests. The agreement rates between conventional tests and the computer
vision model are 85.95\% and 79.69\% for residual plots containing
heteroskedasticity and non-linearity patterns, respectively. These
figures are higher than those calculated for visual tests conducted by
human, indicating that the computer vision model exhibits behaviour more
akin to the best available conventional tests. However, Figure
\ref{fig:conv-mosaic} shows that conventional tests tend to reject
residual plots almost exclusively when the computer vision model does,
but they also reject plots when the computer vision model does not. This
suggests that conventional tests are more sensitive than the computer
vision model.

When plotting the decision against the distance, as illustrated in
Figure \ref{fig:power}, several observations emerge. Firstly, for the
computer vision model, when the distance \(D > 4\), almost all residual
plots are rejected. However, for visual tests conducted by humans, the
threshold is higher, set at \(D > 5\). And conventional tests exhibit
some non-rejection cases for \(D > 3\). Moreover, the computer vision
model tends to have fewer rejected cases than conventional tests when
\(D < 2\). However, visual tests demonstrate the lowest sensitivity to
residual plots with small distances. Therefore, tests based on the
computer vision model are less sensitive to small deviations from model
assumptions than conventional tests but more sensitive to moderate
deviations. Rejection decisions are fitted by logistic regression models
with no intercept terms but an offset equals to
\(\text{log}(0.05/0/95)\). The fitting curves for the computer vision
model fall between those of conventional tests and visual tests for both
non-linearity and heteroskedasticity. There is still potential to refine
the computer vision model to better align its behaviour with visual
tests.
